Supplementary material

### 2021-12-07 analysis/supplementary-materials/supplementary.Rmd
Working through supplementary material

As with Gunnarsrød 6a above, the ^14^C-date to the Stone Age from Gunnarsrød 6b are not related to the Mesolithic inventory on the site [@carrasco2014].

@@

Hesthag C4

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

Lunaveien is a grave cairn site, and the report states that it is unlikely that any of the finds could relate to a settlement predating the construction of the cairn [@reitan2010]. The results have therefore been excluded from the aggregative analysis.

@@

A section of the highway located a few hundred meters to the south was pre-emptively edited in the DTM. This does not appear to be relevant for the simulated sea-level.

@@

@@

@@

@@

@@

@@

>>>>

As with Gunnarsrød 6a above, the ^14^C-date to the Stone Age from Gunnarsrød 6b is not related to the Mesolithic inventory on the site [@carrasco2014].

@@

Typology and radiocarbon date match, and there are no modern disturbances in the vicinity of the site [@viken2018b]. The second ^14^C-date is a reference date of charcoal from a context believed not be anthropogenic.

@@

It was noted in the report that it is uncertain how close to the shoreline Hovland 1 would have been situated, based on the ^14^C-dates [@olsen2013b]. The earliest dates match the lithic inventory. The later date is the only indication of activity from the Early Bronze Age/Early Iron Age. The site is not located far from the highway, which was been edited for the analysis of Hovland 1 and 4, as well as Torstvet (see below).

@@

Hovland 3 is securely dated both typologically and radiometrically [@solheim2013]. It is, however, located just by the highway. This appears to have been successfully removed.

@@

Hovland 4 is well-dated both typologically and with ^14^C-dates to the Mesolithic. It has a similar location to that of Hovland 3 (above), and editing of the highway seems to have been successful.

@@

Hovland 5 is situated more withdrawn from the highway than the other Hovland sites (above). It only has a single ^14^C-date, but this does match the typological indicators of the assemblage [@mansrud2013].

@@

The lithic inventory from Hydal 4 is distinctly Middle Mesolithic in character, and therefore does not match the ^14^C-dates to the Late Mesolithic [@koxvold2017e]. The simulation of the sea-level does not appear impacted by any modern disturbances.

@@

All ^14^C-dates from Krøgenes D1 are believed to be related to use of the site, which has been reused multiple times over a long time-span and which is evidenced in the artefact inventory as well [@reitan2018]. To what degree the different dates are related to different or the same settlement phases is difficult to ascertain, but is here, as with for the rest of the sites, simply based on grouping dates overlapping with 99.7% probability. The site is situated around 100 meter from a road to the south, but this does not appear to impact simulation results, and is therefore left unedited in the DTM.

@@

The oldest date is not seen as relevant for the main occupation on the site, in part also due to the large error [@mansrud2018b]. The site location is similar to that of Krøgenes D1 (above), and did not require any editing.

@@

^14^C-date and lithic inventory at Kvastad A1 correspond with a date to the end of the Early Mesolithic [@stokke2018]. The highway to north of the site is clearly visible on the DTM, but does not impact the simulation results.

@@

The two find areas are predominantly Mesolithic (to the south-west) and Nesolithic (to the north-east) but were treated in combination here as one Mesolithic date is from the Neolithic area [@stokke2018b]. There is also a late Neolithic element in the lithic inventory that is not reflected in the ^14^C-dates. As with Kvastad A1 (above) the highway runs north of the site, clearly visible on the DTM, but does not impact the simulation results.

@@

While there are a few elements of the lithic inventory from Kvastad A4 that could be related to the Stone Age date, it is not related to main visit to the site, which are Early and Middle Mesolithic in character [@darmark2018c]. The highway runs over the site today, but this is not relevant for the shore-line reconstruction done here.

@@

The Stone Age date from Kvastad A9 is not related to the Early Mesolithic assemblage [@darmark2018b]. The highway running by the site does not seem relevant to the simulation results.

@@

Langangen Vestgård 1 is a securely dated site with both ^14^C-dates and typology [@melvold2014]. No modern disturbances appear relevant for the sea-level adjustment.

@@

Typology of the site inventory from Langangen Vestgård 3 match the ^14^C-date [@eggen2014a]. Typological indicators are too coarse to offer any insight on the division into two phases done here for the ^14^C-dates. No editing of the DTM was necessary.

@@

Langangen Vestgård 5. Typology match the ^14^C-dates and a division of two phases, one Late Mesolithic and one Early Neolithic [@reitan2014b]. No editing of the DTM necessary.

@@

Pottery and the lithic inventory of Langanven Vestgård 6 match ^14^C-dates to the Early Neolithic [@reitan2014c]. Nothing in the artefact inventory could be related to the Late Neolithic date. The DTM did not require editing.

@@

Langangen Vestgård 7 is represented by a cooking pit found on one of two terraces scattered with lithics that was not known or originally part of the excavation project  [@reitan2014j]. It was therefore only investigated to a limited degree. However, the material does typologically match the ^14^C-date.

@@

Langemyr is typologically and radiometrically dated to the Middle Mesolithic [@koxvold2018a]. Construction area just south the site was removed. This is the same highway-project and raster editing that was done for some of the Hovland sites above.

@@

Løvås 1. No editing of the raster appeared to be necessary. Interestingly only the Løvås sites were simulated to be situated beneath the sea-level. The typologically secure inventory combined with the many overlapping dates from various features does indicate that the issue is with the displacement curve and not the dating of the site. Similar situation to that on Løvås 2 and 3 (below).

@@

Løvås 2. Similar situation to that on Løvås 1 and 3 (above and below).

@@

Similar situation to that on Løvås 1 and 2 (above).

@@

Lunaveien is a grave cairn site, and the report states that it is unlikely that any of the finds relate to a settlement pre-dating the construction of the cairn [@reitan2010]. The results have therefore been excluded from the aggregative analysis.

@@

Typology of the lithic inventory on Nauen A points to a Late Mesolithic date [@persson2008]. The later date is seen as related to the establishment of a field for grazing or agriculture, which is also supported by dates that fall outside the analysis here. A section of the highway located a few hundred meters to the south was pre-emptively edited in the DTM. This does not appear to be relevant for the simulated sea-level.

Nedre Holtan.

@@

The typological indicators of the lithic inventory from Nauen C match the earliest dates [@persson2008]. As with Nauen A, above, the DTM was edited pre-emptively and does not appear to have been necessary.

@@

Building structures from the Late Neolithic, Bronze Age and Early Iron Age [@gjerpe2008b]. Lithics and pottery finds support this. Site located on what is today a highway which was edited also for Nordby 52 (below), but is not relevant for the sea-level reconstruction for Nordby 1.

@@

Nordby 52 is a rock-shelter tat has been visited throughout prehistory [@gjerpe2008]. Clear typological date to both of the Stone Age periods defined by the ^14^C-dates treated here. The site is located by where the highway runs today (see also Nordby 1 above). Edit appears to have been successful.

@@

Pauler 1 has a distinctly Early Mesolithic artefact inventory that is not related to the ^14^C-dates [@schallerahrberg2012]. No modern disturbances appear to imapact the reconstruction of the sea-level corresponding to the ^14^C-dates.

@@

As with Pauler 1 (above), Pauler 2 has a distinctly Early Mesolithic artefact inventory that is not related to the ^14^C-dates [@nyland2012]. No modern disturbances appear to imapact the reconstruction of the sea-level corresponding to the ^14^C-dates.

@@

Pjonkerød R1. Situated by a quarry today which has been fairly successfully edited. However, as the few negative values across the results are likely to result from the interpolation process and the fact that this is a quite large area, and not issues with the displacement curves or ^14^C-dates, the negative values were forced to zero for the aggregative analsysis.
